---
title: "Calculating heritability from the parent-offspring regression"
output: html_notebook
---

Josef C Uyeda & Stevan J Arnold
(original 2011, revised July 9, 2021)

# Estimating heritability for garter snake litters
## Getting data into R

Get your data into R by following the following steps. Save the entire Exercise 
1.1 directory or download the script and data file. We will use the data file:
 * R_inland_snake_data.txt
If needed, set your working directory by putting the path to the downloaded 
folder you created in the following statement. If you opened this R session and 
script from the Exercise directory, you can use relative paths and skip this step.
Be sure and use the / not the \ spacing convention!

```{r}
library(dplyr)
setwd('./') #You may need to set your working directory to where you downloaded the files
```

Our dataset contains garter snake scale counts for mothers and their offspring. 
Scale counts are good for this sort of comparison because they do not vary with
ontogeny. Why is this important? Why does this simplify estimating heritability 
from a parent-offspring regression?

```{r}
thamnophis <- read.table('R_inland_snake_data.txt',header=TRUE)
thamnophis
```

Here, each family represents a gravid female snake that was caught and all of her 
offspring. The first entry for a given family is the mother's trait 
data. The subsequent entries are her kids. We will use `dplyr` to reorganize 
the dataset into two sets of columns for each family calculating
the mother's and the kids' trait means. Male snakes tend to have higher 
vertebral counts than female snakes. This dataset has already had the effects
of sex removed (which explains why some kids have fractional scale counts).

```{r}
thamnophis <- group_by(thamnophis, family) # Use the variable `family` as a grouping variable
moms <- filter(thamnophis, row_number()==1) # Extract first row for each family
kids <- filter(thamnophis, row_number()!=1) %>% dplyr::summarize(., across(body:post, ~ mean(.x, na.rm=TRUE))) #Extract all but the first row for each family, and summarize by averaging all traits, from `body` to `post`.
families <- left_join(moms, kids, "family", suffix=c(".mom", ".kids")) #Re-join mom and kids
families
```

Now we can estimate heritabilities from a parent-offspring regression. Plot the 
mother and offspring body vertebrae and fit a linear regression. 
Then add a trendline from the linear regression fit. Note that the second line 
will not run without you filling in the appropriate equation.

```{r}
plot(families$body.mom, families$body.kids, pch=19, col="blue", cex=1.5, xlab="Body (Dam)", ylab="Body (Offspring)")
lm1 <- ??? # Fill in the appropriate call here
abline(lm1, lwd=5, lty=2)
summary(lm1)

#Questions:
#What parameter is associated with the heritability? 
#How do we obtain our estimate of the heritability for these data (remember the relatedness)? 
#Is this narrow-sense or broad-sense heritability?
```

Save the results below. Note that the lines marked with `???` will throw an 
error. Your job is to replace with the appropriate equation and variables.

```{r}
h2 <- list() #heritability
Vp <- list() #phenotypic variance
Va <- list() #additive genetic variance

h2$body <- ??? #Remember we only have 1 of the parents. Use the slope from the linear model 
               # by using `lm1$coefficients[2]`
Vp$body <- var(moms$body, na.rm=TRUE) # Could also include offspring, but non-independent samples
Va$body <- ??? # Use equation for heritability
```

Now repeat the analysis for tail (and other variables).

```{r}


```

How much confidence should we have in these estimates? One way to estimate
uncertainty and confidence limits with minimal distributional assumptions is
to conduct a bootstrap.

```{r}
nreps <- 1000 # Number of bootstrap replicates
bootsample <- matrix(NA, nrow=nreps, ncol=6) # Create an empty matrix where we will store results
colnames(bootsample) <- colnames(thamnophis)[3:8] # Set the column names of the empty results matrix
for(i in 1:nreps){
  samp <- sample(1:nrow(families), nrow(families), replace=TRUE) # Key step to a bootstrap: Resample families with replacement
  .newfamilies <- families[samp,] #Use the resampled families to create a new dataset where different families are randomly reweighted in frequency
  h2row <- list() #Stpre results for each trait
  for(j in 3:8){ #iterate over each of the traits
    trait <- colnames(thamnophis)[j] 
    h2row[[trait]] <- 2*unname(lm(.newfamilies[[j+6]] ~ .newfamilies[[j]])$coefficients[2])
  }
  bootsample[i,] <- do.call(c, h2row)
}

```

Now we can plot our resulting distributions:

```{r}

par(mfrow=c(3,2))
for(i in 1:ncol(bootsample)){
  traitname <- colnames(bootsample)[i]
  hist(bootsample[,i], main=traitname, xlim=c(-0.5, 1.5), breaks=nreps/20, xlab="heritability") 
  abline(v=mean(bootsample[,i]), lty=2, lwd=3)
  abline(v=0, lwd=2, col="red")
  abline(v=1, lwd=2, col="red")
}
```

Here, we ignored variation in the sizes of families. How should that affect our 
analyses? What have we learned how to do? This is admittedly a simplistic method 
of estimating heritability and additive genetic variance. We will learn about 
more sophisticated tools in the lecture and lab with Patrick Carter. How might 
our estimates be biased or an inaccurate estimate of the true heritability and 
additive genetic variation? 
